# Lecture 2: Introduction to Causal Inference {#lecture2}

## Slides{.unnumbered}

- 3 Introduction to Causal Inference ([link](https://github.com/svallejovera/regression_ci/blob/main/slide/2%20What%20is%20Causal%20Inference.pptx)) 

## Introduction

We now dive deeper into causal inference and the counterfactual problem. We show why randomized trials solve the counterfactual problem, but also why it remains a challenge when using observational data.

The lecture slides are displayed in full below:

```{r week3slides, echo=FALSE,fig.align = 'center', out.width = "100%", fig.cap = "Slides for 3 Introduction to Causal Inference."}
knitr::include_graphics("slide/3_Introduction_to_Causal_Inference.pdf")
```


```{r, warning = F, message = F}
library(tidyverse) # for wrangling data
library(tidylog) # to know what we are wrangling
```

## Vignette 2.1

Usually, we do not know the data-generating process, but here, we are gods. Let's create a world where taking *treatment A* (e.g., taking a pill) positively *affects Y* (e.g., health) by one unit. Let's run an experiment.

```{r}
df <- data.frame(
  # Simulate baseline health outcomes *without* treatment for 5,000 individuals
  health_no_pill = rnorm(5000),
  # Randomly assign treatment: pill = 1 (treated) or 0 (control)
  pill = sample(c(0, 1), 5000, replace = TRUE)
)

# Visualize the distribution of baseline (no-pill) health outcomes
hist(df$health_no_pill)
```

```{r}
knitr::kable(table(df$pill), format="markdown")
```

Now we can create our counterfactual:

```{r}
df <- df %>%
  # Define the potential outcome under treatment (A = 1):
  # health_w_pill = health_no_pill + 1 implies a constant treatment effect of +1 for everyone.
  mutate(health_w_pill = health_no_pill + 1) # Y(1): the potential outcome if treated

```

Let's look at our counterfactual:

```{r}
# Reshape the two potential outcomes into a long format for easy plotting/comparison
health_w_pill <- cbind.data.frame(df$health_w_pill, "with Pill")
colnames(health_w_pill) <- c("health", "treatment")

health_no_pill <- cbind.data.frame(df$health_no_pill, "without Pill")
colnames(health_no_pill) <- c("health", "treatment")

# Stack the two datasets to compare distributions under treatment vs. no treatment
comparison_y <- rbind.data.frame(health_w_pill, health_no_pill)

comparison_y %>%
  group_by(treatment) %>%
  # Compute group means (used to add mean lines to the plot)
  mutate(mean_health = mean(health)) %>%
  ungroup() %>%
  ggplot(aes(x = health, fill = treatment, color = treatment)) +
  # Plot the distribution of health outcomes under each condition
  geom_density(alpha = 0.5) +
  scale_x_continuous(breaks = scales::pretty_breaks(n = 8)) +
  # Add dashed vertical lines at the mean health for each condition
  geom_vline(
    aes(xintercept = mean_health, color = treatment),
    linetype = "dashed"
  )
```

Now let's give each individual the treatment (either the pill or a placebo):

```{r}
df <- df %>%
  # Construct the observed outcome:
  # - if pill == 1, we observe the treated potential outcome Y(1)
  # - if pill == 0, we observe the untreated potential outcome Y(0)
  mutate(health_obs = ifelse(pill == 1, health_w_pill, health_no_pill))

# Inspect the first 10 rows to verify the observed outcome was assigned correctly
head(df, 10)
```

We can see the average effect of the pill on the treated group (remember from the lecture that the *effect* is, in essence, the *difference* between those who receive the treatment and those who do not):

```{r}
df %>%
  # Compare average observed health outcomes for treated (pill = 1) vs. control (pill = 0)
  group_by(pill) %>%
  summarize(health = mean(health_obs))
```

Or we can plot it:

```{r}
df %>%
  group_by(pill) %>%
  # Compute the mean observed health within treated vs. control groups
  mutate(mean_health_obs = mean(health_obs)) %>%
  ungroup() %>%
  ggplot(aes(x = health_obs, fill = factor(pill), color = factor(pill))) +
  # Plot the distribution of observed outcomes by treatment status
  geom_density(alpha = 0.5) +
  scale_x_continuous(breaks = scales::pretty_breaks(n = 8)) +
  # Add dashed vertical lines at the group means
  geom_vline(
    aes(xintercept = mean_health_obs, color = factor(pill)),
    linetype = "dashed"
  )

```

## Vignette 2.2

Ok... but what happens if we cannot randomize? What if we have observational data, such that...

```{r}
df <- data.frame(income = runif(10000)) %>%
  # Simulate baseline health without treatment (Y(0)).
  # Here, health depends on a random component plus income (so income confounds health).
  mutate(
    health_no_pill = rnorm(10000) + income,
    # Define the treated potential outcome (Y(1)) with a constant +1 treatment effect
    health_w_pill = health_no_pill + 1
  ) %>%
  # Assign treatment non-randomly: only higher-income individuals receive the pill
  mutate(
    pill = income > 0.7,
    # Construct the observed outcome based on treatment assignment
    health_obs = ifelse(pill == 1, health_w_pill, health_no_pill)
  )

# Inspect the first 10 rows
head(df, 10)

```

Let's see what happens now to the estimated mean average 'effect' (remember from the lecture that the *effect* is, in essence, the *difference* between those who receive the treatment and those who do not):

```{r}
df %>%
  group_by(pill) %>%
  summarize(health = mean(health_obs))
```

Oh no! That is more than the true effect of the pill, which we *know* is 1 because we created it. However, if we model this properly (this is an RDD!), then—remember from the lecture that the *effect* is, in essence, the *difference* between those who receive the treatment and those who do not:

```{r}
df %>%
  # Keep observations close to the cutoff (income = 0.7),
  # mimicking the local comparison at the heart of a regression discontinuity design (RDD)
  filter(abs(income - 0.7) < 0.01) %>%
  # Compare mean observed outcomes just below vs. just above the cutoff
  group_by(pill) %>%
  summarize(health = mean(health_obs)) # BOOM!!
```

## Lecture Assignment 

1. Using simulated data, describe (in text) and show (with plots) a plausible situation where a treatment has an effect, but the observed outcome is null. **Tip:** Before committing to an example, remember that a difference-in-differences approach requires two periods (change over time) and two groups (treatment and control).

2. Using a [difference-in-differences approach](#lecture9), show how to estimate the *true* effect from your example.
