# Lecture 2: Introduction to Causal Inference

## Slides{.unnumbered}

- 3 Introduction to Causal Inference ([link](https://github.com/svallejovera/regression_ci/blob/main/slide/2%20What%20is%20Causal%20Inference.pptx)) 

## Introduction

We now dive deeper into causal inference and the counterfactual problem. We show why randomized trials solve the counterfactual problem, but also why it remains a challenge when using observational data.

The lecture slides are displayed in full below:

```{r week3slides, echo=FALSE,fig.align = 'center', out.width = "100%", fig.cap = "Slides for 3 Introduction to Causal Inference."}
knitr::include_graphics("slide/3_Introduction_to_Causal_Inference.pdf")
```


```{r, warning = F, message = F}
library(tidyverse) # for wrangling data
library(tidylog) # to know what we are wrangling
```

## Vignette 2.1

Usually, we do not know the data-generating process, but here, we are gods. Let's create a world where taking *treatment A* (e.g., taking a pill) positively *affects Y* (e.g., health) by one unit. Let's run an experiment.

```{r}
df <- data.frame(health_no_pill= rnorm(5000),
                 # Randomly assign a treatment
                 pill=sample(c(0,1),5000,replace=T))
hist(df$health_no_pill)
```

```{r}
knitr::kable(table(df$pill), format="markdown")
```

Now we can create our counterfactual:

```{r}
df <- df %>%
  mutate(health_w_pill = health_no_pill + 1) # Our Y when A=1 aka our counterfactual
```

Let's look at our counterfactual:

```{r}
health_w_pill <- cbind.data.frame(df$health_w_pill,"with Pill")
colnames(health_w_pill) <- c("health","treatment")
health_no_pill <- cbind.data.frame(df$health_no_pill,"without Pill")
colnames(health_no_pill) <- c("health","treatment")
comparison_y <- rbind.data.frame(health_w_pill,health_no_pill)

comparison_y %>%
  group_by(treatment) %>%
  mutate(mean_health = mean(health)) %>%
  ungroup() %>%
  ggplot(aes(x=health,fill = treatment,color = treatment)) +
  geom_density(alpha = .5) +
  scale_x_continuous(breaks = scales::pretty_breaks(n = 8)) +
  geom_vline(aes(xintercept = mean_health, color = treatment ),
             linetype = "dashed")
```

Now let's give each individual the treatment (either the pill or a placebo):

```{r}
df <- df %>%
  mutate(health_obs = ifelse(pill==1,health_w_pill,health_no_pill))
head(df,10)
```

We can see the average effect of the pill on the treated group (remember from the lecture that the *effect* is, in essence, the *difference* between those who receive the treatment and those who do not):

```{r}
df %>%
  group_by(pill) %>%
  summarize(health = mean(health_obs))
```

Or we can plot it:

```{r}
df %>%
  group_by(pill) %>%
  mutate(mean_health_obs = mean(health_obs)) %>%
  ungroup() %>%
  ggplot(aes(x=health_obs,fill = factor(pill),color = factor(pill))) +
  geom_density(alpha = .5) +
  scale_x_continuous(breaks = scales::pretty_breaks(n = 8)) +
  geom_vline(aes(xintercept = mean_health_obs, color = factor(pill) ),
             linetype = "dashed")

```

## Vignette 2.2

Ok... but what happens if we cannot randomize? What if we have observational data, such that...

```{r}
df <- data.frame(income = runif(10000)) %>%
  # In this case, your health is determined randomly AND by your levels of income
  mutate(health_no_pill = rnorm(10000) + income,
         health_w_pill = health_no_pill + 1) %>%
  # Now we give the pill only to people that have money
  mutate(pill = income > .7,
         health_obs = ifelse(pill==1,health_w_pill,health_no_pill))

head(df,10)
```

Let's see what happens now to the estimated mean average 'effect' (remember from the lecture that the *effect* is, in essence, the *difference* between those who receive the treatment and those who do not):

```{r}
df %>%
  group_by(pill) %>%
  summarize(health = mean(health_obs))
```

Oh no! That is more than the actual effect of the pill, which we *know* is 1 since we created it. However, if we were to properly model (this is an RDD!), then (remember from the lecture that the *effect* is, in essence, the *difference* between those who receive the treatment, and those who do not):

```{r}
df %>%
  filter(abs(income-.7)<.01) %>%
  group_by(pill) %>%
  summarize(health = mean(health_obs)) ## BOOM!!
```


