# Lecture 3: Core Concepts of Experimental Design

## Slides{.unnumbered}

- 3 Introduction to Causal Inference ([link](https://github.com/svallejovera/regression_ci/blob/main/slide/4_Core_Concepts_of_Experimental_Design.pptx)) 

## Introduction

In this lecture we look at a (slightly) more technical understanding of selection bias (to understand the problem with observational data) and the potential outcomes approach (to understand how random assignment solves this problem). 

The lecture slide are displayed in full below:

```{r week4slides, echo=FALSE,fig.align = 'center', out.width = "100%", fig.cap = "Slides for 4 Core Concepts of Experimental Design."}
knitr::include_graphics("slide/4_Core_Concepts_of_Experimental_Design.pdf")
```

# Vignette 3.1: ----

Liebman, Jeffrey B., and Erzo FP Luttmer. "Would people behave differently if they better understood social security? Evidence from a field experiment." American Economic Journal: Economic Policy 7.1 (2015): 275-99.

```{r, warning = F, message = F}
library(tidyverse) # for wrangling data
library(tidylog) # to know what we are wrangling
library(sjPlot) # to plot some models
library(readstata13) # to load .dta files
```

```{r}
exp_data <- read.dta13("Sample_data/Data_WithAgeInconsistency.dta")
```

Our variable of interest (DV, Y) is paid_work_yes: whether the person worked or not. Our treatment variable is treat: whether they got the pamphlet or not. Because this is an experiment, we can compare the averages of the treatment group and the control group to estimate the causal effect. 

```{r}
exp_data %>%
  group_by(treat) %>% 
  summarise(ATE = mean(paid_work_yes, na.rm = T))
```

Same effect as reported by authors!! We can also estimate differences by gender (why?):

```{r}
exp_data %>%
  group_by(female,treat) %>% 
  summarise(ATE = mean(paid_work_yes, na.rm = T))
```

Treatment only has an effect on women... why?


These averages look great and all but how can we be sure if these effects are not only a mere coincidence product of randomization? Let's add some confidence intervals:

```{r}
exp_data$treat <- as.factor(exp_data$treat)
exp_data$female <- as.factor(exp_data$female)

# Full model
model_1 <- lm(paid_work_yes ~ treat, data = exp_data)
summary(model_1)

plot_model(model_1, type = "pred", terms = "treat") +
  theme_minimal() +
  labs(x="Treatment", y="Worked +1 Year")
```

And now by gender;

```{r}
# Model by gender
model_2 <- lm(paid_work_yes ~ female + treat + female*treat, data = exp_data)
summary(model_2)

plot_model(model_2, type = "int") +
  theme_minimal() +
  labs(x="Women", y="Worked +1 Year", color = "Treatment")
```

Cool...

