# Lecture 6: The Multiple Regression Model II

## Slides{.unnumbered}

- 5 The Simple Regression Model ([link](https://github.com/svallejovera/regression_ci/blob/main/slide/7_Multiple_Regression_Model_I.pptx)) 

## Introduction

```{r echo=FALSE}
knitr::opts_chunk$set(echo=TRUE, warning=F, message=T, eval=T)
library(ggpubr)
library(tidyverse)
```

We continue studying the simple regression model. 

```{r week7slides, echo=FALSE,fig.align = 'center', out.width = "100%", fig.cap = "Slides for 7 The Multiple Regression Model."}
knitr::include_graphics("slide/7_Multiple_Regression_Model_I.pdf")
```

## Vignette 6.1

Once again, let's simulate some data.  Maybe we are interested in urban and rural towns (70% are urban) :

```{r}
df <- tibble(urban = sample(c(0,1),500,replace=T,prob=c(.3,.7))) %>%
  ## Urban towns spend, on average, $3 million more on wages than rural towns
  mutate(expen_wages = 3*urban+runif(500,min=0,max=4)) %>%
  ## Urban towns are also have greater incomes (e.g., from taxes), but these are reduced by their high wage expenditures:
  mutate(log_income = 1 + 2*urban - .3*expen_wages + rnorm(500,mean=2)) ## <- Population Eq.
```

Now we can estimate the effect of wage expenditure on income:

```{r}
model_a <- lm(log_income ~ expen_wages, data = df) 
summary(model_a) 
```

Wait what? *(Interpret a log ~ level)*

## Vignette 6.2

Let's see... How can we remove everything from wages that is explained by urban? How can we remove everything from income that is explained by urban?

```{r}
df %>% group_by(urban) %>%
  summarise(income_urb= mean(log_income)) 

df %>% group_by(urban) %>% 
  summarise(expen_wages_urb = mean(expen_wages))
```

The difference between what is explained by urban of income/expendinture (mean) and the observed value of income/expenditure is... 

```{r}
df <- df %>% group_by(urban) %>%
  mutate(log_income_residual = log_income - mean(log_income),
         expen_wages_residual = expen_wages - mean(expen_wages)) %>%
  ungroup()
```

The residual... what is not explained by urban!!

```{r}
model_b <- lm(log_income_residual ~ expen_wages_residual, data = df) 
summary(model_b) ### CLOSER!
```

Let's plot:

```{r}
A <- ggplot(df, aes(x=expen_wages,y=log_income)) +
  geom_point() +
  labs(title = "0. Relation between wages and income. Beta = 0.13") +
  geom_smooth(method = "lm") +
  xlim(c(-3,7)) + ylim(c(-3,6))
A

B <- ggplot(df, aes(x=expen_wages,y=log_income,color = factor(urban))) +
  geom_point() +
  labs(title = "1. Relation between wages and income divided by urban.") +
  xlim(c(-3,7)) + ylim(c(-3,6))
B

C <- ggplot(df, aes(x=expen_wages_residual,y=log_income,color = factor(urban))) +
  geom_point() +
  labs(title = "2. We remove the difference of wages explained by urban.")+
  xlim(c(-3,7)) + ylim(c(-3,6))
C

D <- ggplot(df, aes(x=expen_wages_residual,y=log_income_residual,color = factor(urban))) +
  geom_point() +
  labs(title = "3. We remove the difference of income explained by urban.")+
  xlim(c(-3,7)) + ylim(c(-3,6))
D

E <- ggplot(df, aes(expen_wages_residual,y=log_income_residual)) +
  geom_point() +
  labs(title = "4. We analize what is left. Beta = -0.22") +
  geom_smooth(method = "lm")+
  xlim(c(-3,7)) + ylim(c(-3,6))
E

ggarrange(A,B,C,D,E,
          common.legend = T,
          ncol = 2,
          nrow = 3)
```
