# Lecture 8: The Multiple Regression Model: Interpretation {#lecture8}

## Slides{.unnumbered}

- 9a The Multiple Regression Model: Interpretation ([link](https://github.com/svallejovera/regression_ci/blob/main/slide/9_Interpretation.pptx)) 

- 9b The Multiple Regression Model: F-Tests ([link](https://github.com/svallejovera/regression_ci/blob/main/slide/10_Ftest.pdf)) 

- 9c The Multiple Regression Model: Heteroskedasticity ([link](https://github.com/svallejovera/regression_ci/blob/main/slide/11_Heteroskedasticity.pdf)) 

No fancy code from this Lecture :( ... 

```{r week9slidesA, echo=FALSE,fig.align = 'center', out.width = "100%", fig.cap = "Slides for 9 The Multiple Regression Model: Interpretation."}
knitr::include_graphics("slide/9_Interpretation.pdf")
```

```{r week9slidesB, echo=FALSE,fig.align = 'center', out.width = "100%", fig.cap = "Slides for 9 The Multiple Regression Model: F-Test"}
knitr::include_graphics("slide/10_Ftest.pdf")
```

```{r week9slidesC, echo=FALSE,fig.align = 'center', out.width = "100%", fig.cap = "Slides for 9 The Multiple Regression Model: Heteroskedasticity"}
knitr::include_graphics("slide/11_Heteroskedasticity.pdf")
```

## Lecture Assignment 

1. Using simulated data, explain the difference between statistical significance and substantive significance, as explained by [Achen (1982)](https://ocul-uwo.primo.exlibrisgroup.com/discovery/search?vid=01OCUL_UWO:UWO_DEFAULT&tab=OCULDiscoveryNetwork&search_scope=OCULDiscoveryNetwork&displayMode=full&bulkSize=10&highlight=true&dum=true&query=any,contains,Interpreting%20and%20Using%20Regression&displayField=all&pcAvailabiltyMode=true), Chapter: Sampling Distributions of Regression Coefficients. Note: you can access the chapter through UWO Library. 
