# Lecture 9: Interactions {#lecture9}

## Slides{.unnumbered}

- 10 Interactions ([link](https://github.com/svallejovera/regression_ci/blob/main/slide/12 Interactions.pdf)) 

```{r lectureinteraction, echo=FALSE,fig.align = 'center', out.width = "100%", fig.cap = "Slides for 10 Interactions."}
knitr::include_graphics("slide/12 Interactions.pdf")
```

## Interactions in Linear Models {#interactions}

Interactions let the effect of one variable depend on the level of another. In social science language: *context conditions the relationship*.

We will build intuition with two vignettes:

1. **State-level turnout (OLS):** We’ll also compare **classical** vs **robust** standard errors.
2. **Individual voting (LPM):** We’ll use a linear probability model to keep interpretation simple, and then discuss trade-offs.

---

Start with a model with an interaction:

$$
y = \beta_0 + \beta_1 X + \beta_2 Z + \beta_3 (X\cdot Z) + \varepsilon
$$

Two key interpretations:

1. **Conditional effect of \(X\)** (a.k.a. “simple slope”):
$$
\frac{\partial E[y]}{\partial X} = \beta_1 + \beta_3 Z
$$

2. **Conditional regression line**:
- If \(Z=0\):  \(E[y|X,Z=0] = \beta_0 + \beta_1 X\)
- If \(Z=1\):  \(E[y|X,Z=1] = (\beta_0+\beta_2) + (\beta_1+\beta_3)X\)

So, \(\beta_3\) is literally the **difference in slopes** (when \(Z\) is a dummy).

We’ll plug in real coefficients below.

---

## Vignette A: Turnout and conditional relationships (OLS) {#vignetteA}

```{r}
library(tidyverse)
library(sjPlot)
library(marginaleffects)
library(modelsummary)

data(election_turnout, package = "stevedata")

election_turnout <- election_turnout %>%
  mutate(
    ss = factor(ss),
    trumpw = factor(trumpw),
    region = factor(region),
    gdppercap_ln = log(gdppercap)
  )

summary(election_turnout$turnoutho)
sd(election_turnout$turnoutho)
```

### A.1 Main example (continuous × dummy): education and swing states

Here is the interaction model:

$$
turnoutho_i = \beta_0 + \beta_1 percoled_i + \beta_2 ss_i + \beta_3(percoled_i \cdot ss_i) + \varepsilon_i
$$

```{r}
model_turnout_cd <- lm(
  turnoutho ~ percoled*ss + gdppercap_ln + trumpw,
  data = election_turnout
)

# Side-by-side: classical vs robust SEs
modelsummary(
  list("Classical SEs" = model_turnout_cd,
       "Robust SEs (HC3)" = model_turnout_cd),
  vcov = list("Classical SEs" = NULL,
              "Robust SEs (HC3)" = "HC3"),
  title = "Turnout with Interaction: percoled × swing state",
  coef_rename = c(
    "(Intercept)" = "Intercept",
    "percoled" = "% College Educated",
    "ss1" = "Swing State",
    "percoled:ss1" = "% College × Swing State",
    "gdppercap_ln" = "GDP pc (log)",
    "trumpw1" = "Trump Won"
  ),
  gof_map = c("nobs","r.squared"),
  stars = TRUE,
  notes = list(
    "Column 1: classical OLS standard errors. Column 2: robust (HC3) standard errors.",
    "*** p<0.001, ** p<0.01, * p<0.05"
  )
)
```

From the model:

- When `ss = 0` (not a swing state), the slope of `percoled` is:
  \[
  \text{slope} = \beta_1
  \]
- When `ss = 1` (swing state), the slope of `percoled` is:
  \[
  \text{slope} = \beta_1 + \beta_3
  \]

Let’s compute and display these “simple slopes” directly:

```{r}
# Simple slopes of percoled at ss=0 and ss=1
slopes(model_turnout_cd, variables = "percoled", newdata = datagrid(ss = levels(election_turnout$ss)))
```

Now let’s plot predicted turnout across `percoled` for swing vs non-swing states:

```{r}
plot_model(model_turnout_cd, type = "pred", terms = c("percoled", "ss")) +
  theme_minimal()
```

And a marginal effect plot is another way of saying the same thing:

```{r}
plot_slopes(model_turnout_cd, variables = "percoled", condition = "ss") +
  geom_hline(yintercept = 0, linetype = "dotted") +
  theme_minimal()
```

---

### A.2 Interaction zoo (OLS)

#### (1) Dummy × Dummy: swing states × Trump-win states

Now we interpret \(\beta_3\) as a **difference-in-differences** term:

```{r}
model_turnout_dd <- lm(
  turnoutho ~ ss*trumpw + percoled + gdppercap_ln,
  data = election_turnout
)

modelsummary(
  model_turnout_dd,
  title = "Turnout: swing state × Trump won (Dummy × Dummy)",
  coef_rename = c(
    "(Intercept)" = "Intercept",
    "ss1" = "Swing State",
    "trumpw1" = "Trump Won",
    "ss1:trumpw1" = "Swing × Trump Won",
    "percoled" = "% College Educated",
    "gdppercap_ln" = "GDP pc (log)"
  ),
  gof_map = c("nobs","r.squared"),
  stars = TRUE
)

plot_model(model_turnout_dd, type = "int") +
  theme_minimal()
```

#### (2) Dummy × Categorical: swing states × region

```{r}
model_turnout_dc <- lm(
  turnoutho ~ ss*region + percoled + gdppercap_ln + trumpw,
  data = election_turnout
)

modelsummary(
  model_turnout_dc,
  title = "Turnout: swing state × region (Dummy × Categorical)",
  gof_map = c("nobs","r.squared"),
  stars = TRUE
)

plot_model(model_turnout_dc, type = "pred", terms = c("region","ss")) +
  theme_minimal()
```

#### (3) Continuous × Continuous: education × state wealth

```{r}
model_turnout_cc <- lm(
  turnoutho ~ percoled*gdppercap_ln + ss + trumpw,
  data = election_turnout
)

modelsummary(
  model_turnout_cc,
  title = "Turnout: percoled × log(GDP pc) (Continuous × Continuous)",
  coef_rename = c(
    "(Intercept)" = "Intercept",
    "percoled" = "% College Educated",
    "gdppercap_ln" = "GDP pc (log)",
    "percoled:gdppercap_ln" = "% College × GDP pc (log)",
    "ss1" = "Swing State",
    "trumpw1" = "Trump Won"
  ),
  gof_map = c("nobs","r.squared"),
  stars = TRUE
)

# Visualize: predicted turnout across percoled at representative GDP values
plot_model(model_turnout_cc, type = "pred",
           terms = c("percoled", "gdppercap_ln [7.5, 8.0, 8.5]")) +
  theme_minimal()

# Marginal effect of percoled across GDP
plot_slopes(model_turnout_cc, variables = "percoled", condition = "gdppercap_ln") +
  geom_hline(yintercept = 0, linetype = "dotted") +
  theme_minimal()
```

---

## Vignette B: Interactions in a Linear Probability Model (TV16) {#vignetteB}

```{r}
data(TV16, package = "stevedata")

TV16 <- TV16 %>%
  mutate(
    female = factor(female),
    collegeed = factor(collegeed),
    # Collapse race for cleaner teaching plots:
    race4 = case_when(
      racef == "White" ~ "White",
      racef == "Black" ~ "Black",
      racef == "Hispanic" ~ "Hispanic",
      TRUE ~ "Other"
    ),
    race4 = factor(race4),
    votetrump = as.numeric(votetrump)
  )

table(TV16$votetrump)
```

### Sidebar: why an LPM here?

We could model a binary outcome using logit/probit. But for the purposes of this class, the LPM is extremely convenient:

- Coefficients are in **percentage points** (because \(y\in\{0,1\}\)).
- Interactions are easy to interpret and visualize with straight lines.
- The main statistical downside is **heteroskedasticity** (so robust SEs are usually a good default).
- Another downside is that predicted values can fall outside \([0,1]\). We can check this.

We’ll treat this as a pedagogical model and focus on interpretation.

---

### B.1 Main example (continuous × dummy): ideology × college education

```{r}
model_tv_cd <- lm(
  votetrump ~ ideo*collegeed + female + race4,
  data = TV16
)

modelsummary(
  list("Classical SEs" = model_tv_cd,
       "Robust SEs (HC3)" = model_tv_cd),
  vcov = list("Classical SEs" = NULL,
              "Robust SEs (HC3)" = "HC3"),
  title = "LPM: Trump vote with interaction (ideo × college)",
  stars = TRUE,
  notes = list(
    "Binary outcome estimated by OLS (Linear Probability Model).",
    "Column 2 uses robust (HC3) standard errors."
  )
)
```

- When `collegeed = 0`, the marginal effect of ideology is \(\beta_1\).
- When `collegeed = 1`, the marginal effect is \(\beta_1 + \beta_3\).
- Because \(y\) is 0/1, interpret these as **percentage point changes in Pr(Trump vote)**.

```{r}
plot_slopes(model_tv_cd, variables = "ideo", condition = "collegeed") +
  geom_hline(yintercept = 0, linetype = "dotted") +
  theme_minimal()

plot_model(model_tv_cd, type = "pred", terms = c("ideo", "collegeed")) +
  theme_minimal() +
  scale_y_continuous(labels = scales::percent)
```

Quick check: do predicted values leave [0,1]?

```{r}
pred_check <- predictions(model_tv_cd) %>%
  summarize(min_pred = min(estimate), max_pred = max(estimate))
pred_check
```

---

### B.2 Interaction zoo (LPM)

#### (1) Dummy × Dummy: female × college

```{r}
model_tv_dd <- lm(
  votetrump ~ female*collegeed + ideo + race4,
  data = TV16
)

modelsummary(
  model_tv_dd,
  title = "LPM: female × college (Dummy × Dummy)",
  stars = TRUE
)

plot_model(model_tv_dd, type = "int") +
  theme_minimal() +
  scale_y_continuous(labels = scales::percent)
```

#### (2) Dummy × Categorical: college × race4 (collapsed)

```{r}
model_tv_dc <- lm(
  votetrump ~ collegeed*race4 + female + ideo,
  data = TV16
)

modelsummary(
  model_tv_dc,
  title = "LPM: college × race4 (Dummy × Categorical)",
  stars = TRUE
)

plot_model(model_tv_dc, type = "pred", terms = c("race4", "collegeed")) +
  theme_minimal() +
  scale_y_continuous(labels = scales::percent)
```

#### (3) Continuous × Continuous: ideology × racial resentment proxy (lemprac)

```{r}
model_tv_cc <- lm(
  votetrump ~ ideo*lemprac + female + collegeed + race4,
  data = TV16
)

modelsummary(
  model_tv_cc,
  title = "LPM: ideo × lemprac (Continuous × Continuous)",
  stars = TRUE
)

plot_model(model_tv_cc, type = "pred", terms = c("ideo", "lemprac [2,4,6]")) +
  theme_minimal() +
  scale_y_continuous(labels = scales::percent)

plot_slopes(model_tv_cc, variables = "ideo", condition = "lemprac") +
  geom_hline(yintercept = 0, linetype = "dotted") +
  theme_minimal()
```

---

## Lecture Assignment 

1. In the turnout model (A.1), write the fitted line for `ss = 0` and `ss = 1`. Identify the slope in each case.
2. In the TV16 model (B.1), interpret the interaction sign: does ideology matter *more* or *less* among college-educated respondents?
3. Pick two values of the continuous variable (e.g., `ideo=2` and `ideo=4`). Compute the predicted probability for college vs non-college using the estimated coefficients and compare.
