# Lecture 10: Difference-in-Difference {#lecture10}

## Slides{.unnumbered}

- 10 Difference-in-Difference ([link](https://github.com/svallejovera/regression_ci/blob/main/slide/13_Difference_in_Difference.pptx)) 

## In-class example 

Below is the example code from class lectures.

```{r in_class_example, eval=FALSE}
## Experiments and Causal Inference
## Week 11: Difference in difference II

library(tidyverse)
library(modelsummary)
library(fixest)
library(readr)

## Vignette 11.1: Difference in Difference Example:

# Data from the interwebs:
od_data ="https://raw.githubusercontent.com/NickCH-K/causalbook/main/DifferenceinDifferences/organ_donation.csv"

od <- read_csv(url(od_data))

# Treatment
od <- od %>%
  mutate(Treatment = State == 'California' & 
           Quarter %in% c('Q32011','Q42011','Q12012'),
         After_Treatment = Quarter %in% c('Q32011','Q42011','Q12012'),
         California = State == 'California')

# Feols = Fixed Effects OLS... Everything afther the | 
# are the variable for which I want to do fixed effects:
# feols also allows us to cluster the errors around the first fixed effect

clfe <- feols(Rate ~ Treatment | State + Quarter,
              data = od)

# Alternative II:
clfe_2 <-lm(Rate ~ Treatment + factor(State) + factor (Quarter),
              data = od)

# Alternative III:
clfe_3 <-lm(Rate ~ After_Treatment + California + 
              After_Treatment * California,
            data = od)

msummary(list(clfe,clfe_2,clfe_3), stars = c('*' = .1, '**' = .05, '***' = .01),
         vcov = list(~State,"robust","robust"),
         coef_omit = "factor", notes = "Fixed effects coefficient in Model 2 are omitted.",
         gof_map = c('nobs',"adj.r.squared",'FE: State','FE: Quarter','vcov.type'))
```


## Difference-in-Differences (DiD) {#did}

Difference-in-differences is a way to estimate a treatment effect when we have:

- a **treated** group and a **comparison** group, and
- outcomes observed **before** and **after** treatment begins.

The core intuition is simple:

> **Use the comparison group to subtract off whatever would have happened over time anyway.**

We’ll do this twice:
1. A clean two-period DiD with real data (**injury.csv**).
2. A short multi-period example (so we can talk about *event study* plots and parallel trends checks).



### The math spine (two-by-two DiD)

Let $T\in\{0,1\}$ indicate treatment group and $Post\in\{0,1\}$ indicate the post period.

The DiD estimand is:

$$
\widehat{\tau}_{DID}=(\bar Y_{T=1,Post=1}-\bar Y_{T=1,Post=0}) - (\bar Y_{T=0,Post=1}-\bar Y_{T=0,Post=0})
$$

That is: *change in treated* (i.e., the difference before and after in the treatment group) minus (i.e., the difference) *change in control* (i.e. the difference before and after in the control group). Thus, a difference-in-difference.



### DiD as a regression interaction

The “canonical” DiD regression is:

$$
Y = \beta_0 + \beta_1 T + \beta_2 Post + \beta_3(T\cdot Post) + \varepsilon
$$

Interpretation cheat-sheet (these are fitted cell means):

- **Control, pre:** $\beta_0$
- **Treated, pre:** $\beta_0+\beta_1$
- **Control, post:** $\beta_0+\beta_2$
- **Treated, post:** $\beta_0+\beta_1+\beta_2+\beta_3$

So $\beta_3$ is the DiD estimate.


Note: the following two examples were borrowed from [Andrew Heiss' course](https://evalsp21.classes.andrewheiss.com/example/diff-in-diff/) on program evaluation.

## Vignette A: Workers’ compensation injury durations (two-period DiD)

We’ll use `injury.csv`. In this dataset:

- `ky` indicates the treated group (Kentucky = 1; Michigan = 0)
- `afchnge` indicates the post period (after the policy change = 1)
- `ldurat` is log duration (a nice, well-behaved outcome)

```{r}
library(tidyverse)
library(modelsummary)
library(marginaleffects)
library(sjPlot)

injury <- read_csv("Sample_data/injury.csv", show_col_types = FALSE)

injury <- injury %>%
  mutate(
    ky = as.integer(ky),
    mi = as.integer(mi),
    afchnge = as.integer(afchnge),
    treated = ky,
    post = afchnge,
    did = treated * post
  )

injury %>% count(treated, post)
summary(injury$ldurat)
```

### A.1 Start with a picture: group means pre vs post

```{r}
means_2x2 <- injury %>%
  group_by(treated, post) %>%
  summarize(
    mean_ldurat = mean(ldurat, na.rm = TRUE),
    n = n(),
    .groups = "drop"
  )

means_2x2
```

Plot those means:

```{r}
means_2x2 %>%
  mutate(
    group = if_else(treated == 1, "Treated (KY)", "Control (MI)"),
    period = if_else(post == 1, "Post", "Pre")
  ) %>%
  ggplot(aes(x = period, y = mean_ldurat, group = group)) +
  geom_point() +
  geom_line() +
  labs(x = NULL, y = "Mean log(duration)") +
  theme_minimal()
```

### A.2 Compute DiD “by hand” from the four means

```{r}
m_t_pre  <- means_2x2 %>% filter(treated==1, post==0) %>% pull(mean_ldurat)
m_t_post <- means_2x2 %>% filter(treated==1, post==1) %>% pull(mean_ldurat)
m_c_pre  <- means_2x2 %>% filter(treated==0, post==0) %>% pull(mean_ldurat)
m_c_post <- means_2x2 %>% filter(treated==0, post==1) %>% pull(mean_ldurat)

did_by_hand <- (m_t_post - m_t_pre) - (m_c_post - m_c_pre)
did_by_hand
```

Because the outcome is logged, it’s also helpful to translate the DiD into a percent change:

```{r}
100 * (exp(did_by_hand) - 1)
```

### A.3 Estimate the same thing with the canonical regression

```{r}
m_did_basic <- lm(ldurat ~ treated * post, data = injury)

modelsummary(
  list("DiD (basic)" = m_did_basic,
       "DiD (basic, robust HC3)" = m_did_basic),
  vcov = list("DiD (basic)" = NULL,
              "DiD (basic, robust HC3)" = "HC3"),
  title = "Two-period DiD as an interaction (treated × post)",
  stars = TRUE,
  notes = list(
    "Robust SEs are useful here because the two-by-two setup often has heteroskedasticity.",
    "Clustering by state is not sensible here because there are only 2 states (2 clusters)."
  )
)
```

#### Plug-in: reconstruct the fitted 2×2 cell means from coefficients

```{r}
b <- coef(m_did_basic)

cell_control_pre  <- b["(Intercept)"]
cell_treated_pre  <- b["(Intercept)"] + b["treated"]
cell_control_post <- b["(Intercept)"] + b["post"]
cell_treated_post <- b["(Intercept)"] + b["treated"] + b["post"] + b["treated:post"]

tibble(
  cell = c("Control, Pre", "Treated, Pre", "Control, Post", "Treated, Post"),
  fitted = c(cell_control_pre, cell_treated_pre, cell_control_post, cell_treated_post)
)
```

Check: the DiD estimate is $\hat\beta_3$, and it should match the by-hand estimate.

```{r}
coef(m_did_basic)["treated:post"]
did_by_hand
```

### A.4 Add covariates (precision, not magic)

Adding controls can improve precision and adjust for observable differences, but the identifying assumption is still **parallel trends**.

```{r}
m_did_controls <- lm(
  ldurat ~ treated*post + male + lage + married + lprewage +
    head + neck + upextr + trunk + lowback + lowextr +
    manuf + construc,
  data = injury
)

modelsummary(
  list("DiD (basic, HC3)" = m_did_basic,
       "DiD + controls (HC3)" = m_did_controls),
  vcov = list("DiD (basic, HC3)" = "HC3",
              "DiD + controls (HC3)" = "HC3"),
  title = "DiD with and without controls (robust SEs)",
  stars = TRUE
)
```

### A.5 Interpreting $\beta_3$ in log points (a quick translation)

If $\hat\beta_3 = -0.06$, that is about a:

$$
100\times(\exp(-0.06)-1) \approx -5.8\%
$$

Let’s compute that translation for the estimated DiD coefficient:

```{r}
b3 <- coef(m_did_controls)["treated:post"]
b3
100*(exp(b3)-1)
```



## Vignette B: A multi-period DiD (for event study intuition)

Two-period DiD is great for the core idea, but it can’t show you pre-trends.

So here’s a quick multi-period example (simulated) where we can:
- plot treated vs control across many periods,
- estimate a two-way fixed effects model,
- and draw an event study plot.

```{r}
library(fixest)

set.seed(123)

n_units <- 100
n_t <- 8
treat_time <- 5

panel <- tidyr::expand_grid(
  id = 1:n_units,
  t = 1:n_t
) %>%
  mutate(
    treated = if_else(id <= n_units/2, 1L, 0L),
    post = if_else(t >= treat_time, 1L, 0L),
    alpha_i = rep(rnorm(n_units, 0, 1), each = n_t),
    gamma_t = rep(seq(-0.3, 0.3, length.out = n_t), times = n_units),
    tau = if_else(treated==1 & post==1, -0.25 - 0.05*(t - treat_time), 0),
    y = 2 + alpha_i + gamma_t + tau + rnorm(n_units*n_t, 0, 0.5)
  )

panel %>%
  group_by(t, treated) %>%
  summarize(ybar = mean(y), .groups="drop") %>%
  mutate(group = if_else(treated==1, "Treated", "Control")) %>%
  ggplot(aes(x = t, y = ybar, group = group)) +
  geom_line() +
  geom_point() +
  geom_vline(xintercept = treat_time, linetype = "dotted") +
  labs(x = "Time", y = "Average outcome", caption = "Dotted line = treatment begins") +
  theme_minimal()
```

### B.1 Two-way FE DiD (TWFE)

```{r}
m_twfe <- feols(y ~ treated*post | id + t, data = panel, cluster = ~id)
etable(m_twfe)
```

### B.2 Event study: leads and lags

Create an adoption time variable (`g`) and estimate event-time effects with `sunab()`.

```{r}
panel <- panel %>%
  mutate(g = if_else(treated==1, treat_time, 0L))  # 0 = never-treated

m_es <- feols(y ~ sunab(g, t) | id + t, data = panel, cluster = ~id)

iplot(m_es, ref.line = 0)
```

Interpretation:
- Pre-treatment coefficients (leads) near 0 → consistent with parallel trends.
- Post-treatment coefficients (lags) show dynamics.



