# Lecture 11: Regression Discontinuity {#lecture11}

## Slides{.unnumbered}

- 11 Difference-in-Difference ([link](https://github.com/svallejovera/regression_ci/blob/main/slide/14_RDD.pdf)) 

```{r setup, echo=FALSE}
knitr::opts_chunk$set(echo=TRUE, warning=FALSE, message=FALSE, eval=TRUE)
```

## Introduction

RDD is what you do when treatment assignment is driven by a **threshold rule**.

- Running variable: `entrance_exam`
- Cutoff: \(c = 70\)
- Treatment (sharp): \(D = 1[\text{entrance\_exam} \ge 70]\)
- Outcome: `exit_exam`

The key idea is **local comparison**:

> Compare units just below 70 to units just above 70.



### The math spine

In a **sharp** RDD, the estimand is the jump in the conditional mean at the cutoff:

$$
\tau_{RDD}=
\lim_{x \downarrow 70} \mathbb{E}[Y \mid X = x]-
\lim_{x \uparrow 70} \mathbb{E}[Y \mid X = x]
$$

Interpretation: \(\tau_{RDD}\) is the treatment effect **right at the cutoff** (not necessarily far away from 70).



## Vignette A: Tutoring program (sharp RDD at 70)

In this hypothetical example, students take an entrance exam at the beginning of a school year. Those who score 70 or below are automatically enrolled in a free tutoring program and receive assistance throughout the year. At the end of the school year, students take a final test, or exit exam (with a maximum of 100 points) to measure how much they learned overall. Remember, this is a hypothetical example and tests like this don’t really exist, but just go with it.

```{r}
library(tidyverse)
library(modelsummary)
library(rdrobust)
library(sjPlot)
library(marginaleffects)

tut <- read_csv("Sample_data/tutoring_program.csv", show_col_types = FALSE) %>%
  mutate(
    cutoff = 70,
    x = entrance_exam,
    y = exit_exam,
    x_c = x - cutoff,
    treat = as.integer(x >= cutoff)
  )

tut %>% summarize(min_x = min(x, na.rm=TRUE), max_x = max(x, na.rm=TRUE))
tut %>% count(treat)
```

### A.1 Picture first: binned means around the cutoff

Before we estimate anything, we want to *see* the discontinuity.

A simple way: bin the running variable and plot mean outcomes in each bin.

```{r}
## Pick a bin width that gives a readable plot
bin_width <- 2

binned <- tut %>%
  mutate(bin = floor((x - min(x, na.rm=TRUE)) / bin_width)) %>%
  group_by(bin) %>%
  summarize(
    x_bin = mean(x, na.rm=TRUE),
    y_bin = mean(y, na.rm=TRUE),
    .groups = "drop"
  )

ggplot(binned, aes(x = x_bin, y = y_bin)) +
  geom_point(alpha = 0.6) +
  geom_line(alpha = 0.6) +
  geom_vline(xintercept = 70, linetype = "dashed") +
  labs(x = "Entrance exam (running variable)", y = "Mean exit exam (binned)",
       caption = "Dashed line = cutoff at 70") +
  theme_minimal()
```

That plot is *suggestive*, but it is not an estimator. Now we estimate the jump formally.



### A.2 Estimate the discontinuity with `rdrobust()`

`rdrobust()` focuses on observations near the cutoff and uses local polynomial regression.

We'll estimate two versions (same idea, slightly different choices):

1. Automatic bandwidth choice (a standard default workflow).
2. A fixed bandwidth and linear fit (so students see what changes when we change the “window”).

```{r}
m1 <- rdrobust(y = tut$y, x = tut$x, c = 70,
               kernel = "triangular")

m2 <- rdrobust(y = tut$y, x = tut$x, c = 70,
               kernel = "triangular", h = 10, p = 1)

summary(m1)
summary(m2)
```

#### Interpretation

The key quantity is the estimated jump at the cutoff (the treatment effect at \(x=70\)).

A plug-in interpretation sounds like:

> “Crossing the cutoff (and getting tutoring) changes exit exam scores by about **X points** for students right around an entrance score of 70.”



### A.3 The built-in RDD plot (`rdplot()`)

This is a convenient diagnostic + presentation plot. It bins the data and overlays the fitted functions on each side.

```{r}
rdplot(tut$y, tut$x,
       c = 70,
       x.label = "Entrance exam",
       y.label = "Exit exam",
       binselect = "qsmvpr")

rdplot(tut$y, tut$x,
       c = 70,
       p = 1,
       x.label = "Entrance exam",
       y.label = "Exit exam",
       binselect = "qsmvpr")
```



### A.4 “Same plot, but prettier” (rebuild in ggplot)

Sometimes you want the rdplot bins, but in your own plotting style.

`rdplot()` returns the binned means it used. We can grab those and build a custom plot.

```{r}
rd_obj <- rdplot(tut$y, tut$x,
                 c = 70,
                 p = 1,
                 x.label = "Entrance exam",
                 y.label = "Exit exam",
                 binselect = "qsmvpr")

rd_bins <- tibble(
  x_bin = rd_obj$vars_bins$rdplot_mean_bin,
  y_bin = rd_obj$vars_bins$rdplot_mean_y
) %>%
  mutate(side = if_else(x_bin >= 70, "Above cutoff", "Below cutoff"))

ggplot(rd_bins, aes(x = x_bin, y = y_bin)) +
  geom_point(alpha = 0.5) +
  geom_smooth(method = "lm", se = FALSE) +
  geom_vline(xintercept = 70, linetype = "dashed") +
  labs(x = "Entrance exam", y = "Exit exam",
       caption = "Points are rdplot binned means; lines are separate linear fits") +
  theme_minimal()
```



### A.5 Sensitivity: bandwidth and polynomial order

RDD estimates are *local*—so choices about how local (bandwidth) matter.

We’ll estimate a small grid and put it in a compact table.

```{r}
specs <- tribble(
  ~h, ~p,
  5,  1,
  10, 1,
  15, 1,
  10, 2
)

sens <- specs %>%
  mutate(
    fit = pmap(list(h, p), \(h, p) rdrobust(y=tut$y, x=tut$x, c=70, h=h, p=p, kernel="triangular")),
    tau = map_dbl(fit, \(m) m$coef[1]),
    se  = map_dbl(fit, \(m) m$se[1]),
    lo  = tau - 1.96*se,
    hi  = tau + 1.96*se
  )

sens %>%
  mutate(spec = paste0("h=", h, ", p=", p)) %>%
  select(spec, tau, se, lo, hi)
```

Interpretation note:

- If the estimate changes wildly across reasonable bandwidths, that’s a warning sign.
- Usually we want to see the estimate remain *qualitatively stable* (direction + rough magnitude).



### A.6 Credibility checks (what we look for)

#### (1) Sorting/manipulation near the cutoff (visual check)

In a sharp RDD, a big red flag is **bunching** right above 70 (people gaming the entrance score).

A simple visual check:

```{r}
tut %>%
  ggplot(aes(x = x)) +
  geom_histogram(bins = 40) +
  geom_vline(xintercept = 70, linetype = "dashed") +
  labs(x = "Entrance exam", y = "Count",
       caption = "Look for unusual bunching right above 70") +
  theme_minimal()
```

#### (2) Covariate balance near the cutoff (if we have covariates)

If the dataset contains pre-treatment covariates, we can run the same RD logic using each covariate as an outcome.
We *want* covariates to be smooth at 70.

```{r}
covars <- setdiff(names(tut), c("entrance_exam","exit_exam","cutoff","x","y","x_c","treat"))

covars
```

If you see reasonable pre-treatment covariates here (e.g., demographics), pick one or two and run:

```{r eval=FALSE}
## Example (turn eval=TRUE once you choose real covariates):
## rdrobust(y = tut$some_covariate, x = tut$x, c = 70, kernel = "triangular") |> summary()
```



## Vignette B: A second RDD example (simulated, so we know the truth)

The tutoring example is “real-data messy.” A simulated example is useful because:
- we *know* whether there is a true jump,
- we can see how bandwidth and functional form interact.

```{r}
set.seed(123)

n <- 2000
cutoff <- 0

sim <- tibble(
  x = runif(n, -2, 2),
  treat = as.integer(x >= cutoff),
  ## Smooth relationship + a true jump at 0
  y = 50 + 6*x - 2*x^2 + 4*treat + rnorm(n, 0, 3)
)

## Binned plot
sim_bins <- sim %>%
  mutate(bin = cut(x, breaks = seq(-2, 2, by = 0.1), include.lowest = TRUE)) %>%
  group_by(bin) %>%
  summarize(x_bin = mean(x), y_bin = mean(y), .groups="drop")

ggplot(sim_bins, aes(x = x_bin, y = y_bin)) +
  geom_point(alpha = 0.6) +
  geom_line(alpha = 0.6) +
  geom_vline(xintercept = 0, linetype = "dashed") +
  labs(x = "Running variable (centered)", y = "Outcome (binned means)",
       caption = "True jump at 0 is +4") +
  theme_minimal()

## rdrobust estimate
sim_m <- rdrobust(y = sim$y, x = sim$x, c = 0, kernel = "triangular")
summary(sim_m)
```

Interpretation:
- Because we built the data with a true jump of +4, a good RDD estimate (with enough data near the cutoff) should recover something close.

## Exercises 

1. In the tutoring example, identify \(X\), \(c\), \(D\), and \(Y\).
2. What does the RDD estimate mean in words? (Be explicit: “for whom” is this causal effect?)
3. Compare two bandwidths (e.g., \(h=5\) vs \(h=15\)). What changed in the estimate and the uncertainty?
4. What pattern in the histogram around 70 would worry you about manipulation?

